The transboundary nature of the world’s exploited marine species

### 2020-12-14 Scripts/Manuscript_Draft.Rmd
Villy's and Tay's comments

I overlaid the known distribution of 938 commercially valuable marine species responsible for an average of 96.5% of global EEZ catches between 2005 and 2014, and 280 EEZs of 198 coastal countries (See Methods). I define a ‘stock’ unit as a species in an EEZ, instead of a genetically or morphological distinct unit [@Nelson:2016kk], due to the lack of such biological information being available for almost all the species included in this study [@Teh:2015gd]. While I acknowledge that species could have multiple stocks within an EEZ, many fisheries within a country or EEZ are managed at the species instead of stock level (or even as groups containing multiple species). For example, shrimp (e.g., *Litopenaeus stylirostris* or *Farfantepenaeus californiensis*) along the Pacific coast of Mexico [@DiarioOficialdelaFederacionDOF:2018wt] and hammerhead shark (*Sphyrna zygaena*) in Peru [@MAP:2017uf] are managed at the species level, yet include multiple populations exploited by each country’s fisheries. Moreover, recent research shows connectivity across fish stocks through larval dispersal [@Ramesh:2019va] and adult migration [@Levin:2018in; @Popova:2019fs; @Dunn:2019ca], although considerable level of uncertainty exists at different life stages [@Kaplan:2016hs; @Archambault:2016ie]. For this analysis I only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g., I did not include the high seas). I rely on multiple data sources including occurrence, distribution models and catch data, and only consider a species to be present in a grid cell if all data sources showed positive occurrence (see Methods).

@@

The previous models combine observational data with a series of biotic and abiotic data to determine the probability that a species will be found in a given space at a given time. However, this does not mean that the species in question will actually be there. While the models do use approaches to double-check species occurrences (e.g., ENM-Nereus uses four different species distribution algorithms and SDM-SAU undertakes validation by means of other models), I used a fourth data set to corroborate the models’ outputs. The Sea Around Us estimates total reconstructed catches - catches based on all publicly available information sources and including discards, as well as unreported and illegal catches that are not included in available FAO data - for each country. Catches are also spatially allocated on a $0.5^{\circ}$ x $0.5^{\circ}$ latitude longitude grid [@Zeller2016]. Roughly, the Sea Around Us method consist of the following steps. First, it takes each country's officially reported catch data (e.g., National, FAO or RFMO). Secondly, it uses literature (e.g., peer review, grey literature) to identify missing components (e.g., species, gears) and sources of alternative information for missing components. It then derives country estimates for missing data and creates time series interpolation. Finally, the estimated and official data are aggregated, making up the total reconstructed catch data (see Pauly *et al* [-@Pauly:2019bf] and Zeller *et al.* [-@Zeller2016] for catch reconstruction and spatial allocation details). I used the Sea Around Us catch reconstruction database from 2005 to 2014 as the fourth dataset to estimate transboundary species and to estimate their catch contribution within EEZs. I selected this time frame to investigate the recent (our time frame includes the last decade of available data) contribution of transboundary fisheries to catches and revenue from fisheries and to reduce the uncertainty embedded in the reconstruction process (see Key Uncertainties). Note that in call cases I report the average catch from 2005 to 2014.

@@

Rules <- c(
  "(Year of Catch > Year Post Max. Min. & Post Max Min Catch < (Max Catch*0.10)) & (Catch > (Max Catch*0.10) & Catch < (Max Catch*0.50)) or Year of Catch < Year of Max. Catch & Catch <= (Max Catch*0.50) or Year of Max Catch = Last Year of data)", #Rebuilding or #Developing
  "Catch > (Max Catch*0.50)", #Max Expploitation
  "Year of Catch > Year of Max. Catch & (Catch > (Max. Catch*0.10) & Catch < (Max. Catch*0.50) or Catch < Max. Catch*0.10", #Over explpoited or collapsed
  "Non of the above rules applied"
)

@@

I have identified four key uncertainties in the method utilized that may affect the estimation of transboundary species. Firstly, as I ran the analysis at the species level due to limited spatial-specific data on species sub populations (stocks), I are not able to identify transboundary stocks within EEZs. While I acknowledge that a species could have multiple stocks within an EEZ, many fisheries within a country or EEZ are managed at the species instead of stock level. Also, recent research suggests strong connectivity between stocks, even when separated by thousands of kilometers [@Ramesh:2019va] providing additional ecological ground for our analysis [@Popova:2019fs; @Dunn:2019ca]. However, it is important to acknowledge that there is considerable uncertainty associated in determining levels of connectivity across time and space for marine populations at different life stages, from larvae [@Kaplan:2016hs] to adults [@Archambault:2016ie]. This is of special concern, but-not-limited-to [@Archambault:2016ie], highly migratory species whose ranges span multiple jurisdictions and the high seas, such as tunas [@Moore:2020cd], billfishes [@Sepulveda:2019hq] and sharks [@Vandeperre:2014jk] challenging management decisions based on meta-populations [@Moore:2020cd]. Here, I limited the definition of “transboundary” to include species spanning only adjacent countries (e.g., the US and Canada), excluding countries that were separated by another nation (e.g., Canada and Mexico) and/or the high seas (e.g., Canada and Spain). Consequently, our results likely provide a conservative estimate of transboundary species, as I did not cover all marine taxa in the world [@Reygondeau:2019uh] and only analyzed species present in all four data sources. Nevertheless, our results are representative of a substantial proportion of the world's marine catches and revenue from economically important marine fisheries. Secondly, the predicted species' distribution is affected by the uncertainties of the environmental data and models used for such predictions. Structural differences within Earth System Models result in variations of oceanic conditions, which undoubtedly affect the ENM-Nereus. I averaged results from the three models in an effort to capture the structural variation across models. Natural climate variability is a major driver of marine species' distributions, potentially removing a species from an EEZ for long periods (e.g., Anchovies and Sardines "regimes" in the Eastern Pacific are strongly influenced by water temperature decal oscillations [@Chavez:2003jr]. Thus, a species’ distribution can potentially be reduced or shifted in such a way that it only covers one EEZ until oceanic conditions change again and the associated species' distribution expands. To account for such climate variability, I derived the ENM-Nereus results as an average of oceanic conditions between 1970 and 2000. This is not an issue for the SDM-SAU as it does not directly require environmental variables [@Pauly:2016uz]. Thirdly, I assumed that, if the Sea Around Us reconstructed data recorded a species as caught in any given grid cell, then the species was present within that grid cell. While catch data are not exempt of uncertainty, in most cases, differences between the Sea Around Us and the FAO self-reported data are smaller towards the end of the time series [@Pauly:2019bf]. Thus, I limited the catch data reference period in our analyses to between 2005 and 2014, the last ten years of data available. Likewise, the spatial allocation of the catch is subjected to imprecision, mainly produced by differences in the spatial scale of the original data and the method employed by the Sea Around Us. Finally, this study’s results only considered species for which all datasets agreed on presence and had and Area Index of at least 25% (e.g., the species shared distribution was at least 25% in each EEZ). Therefore, again, our approach presents a relatively conservative estimate of the number of transboundary species in the world. Using a smaller Area Index will result in more transboundary species (Figure \@ref(fig:ffvFigureS2)).

@@

```{r ffvFigure2, eval = T, fig.width=6, fig.height=4, echo=FALSE, fig.pos= "H", fig.cap="Weighted benefits of transboundary species by km2 and UN sub-region. a) Revenue in thousand 2019 USD. b) Catch in tonnes. Points = sub-region mean \u00B1 s.d. by country."}

@@

I determined the catch trend of each species within each EEZ, classifying them as increasing (Category A), constant (Category B) or decreasing (Category C) (Figures \@ref(fig:ffvFigure3); \@ref(fig:ffvFigureS2)). While previous studies have demonstrated that catch trends can be used to infer whether a stock is healthy, re-building, over-exploited or collapsed [@Kleisner:2011wn], several factors can influence stock status. Our intention here is to examine where the catch trends of transboundary species differ from non-transboundary species between 2005 and 2014 relative to historic catch since 1951 (see Methods). I found significant differences for all catch trend categories for transboundary species (one-way ANOVA, DF(2,459) = 47.94, p < 0.001, $\alpha$ = 0.05; Tukey's post hoc test p $\leq$ 0.001), and no significant differences in catch trend categories for discrete species (one-way ANOVA, DF(2,106) = 1.885, p = 0.157; $\alpha$ = 0.05). I also found significant differences in catch trends when directly comparing transboundary to discrete species categories (MANOVA, DF(2,459) = 19.001, p < 0.001). Overall, transboundary species only targeted by one country are generally less likely to have a decreasing catch trend compared to instances where the shared species is fished by neighboring countries (Table \@ref(tab:ffvTableS2)).

@@

Transboundary fisheries are important to a number of countries with notorious fisheries-related conflicts, including Canada, the USA, the European Union (EU) and Russia [@Spijkers:2019cz]. For example, since 2007, the EU, Norway, Iceland, and the Faroe Islands (Denmark) have been at odds over the size and relative allocation of the total allowable catch for Atlantic mackerel (*Scomber scombrus*) due to the species' climate-driven shift in distribution [@Spijkers:2017ij]. Atlantic Mackerel is a transboundary species that straddles into the high seas. Among the countries involved in the 2007 fisheries dispute, Atlantic mackerel contributed an annual average (between 2005 and 2014) of 598.4 thousand tonnes (8.19%) in total catch and USD 850.2 million (7.17%) in total fishing revenue. Climate change is expected to continue changing the distribution and shared proportion of fish stocks among countries, resulting in the emergence of new transboundary species [@Pinsky:2018cb, Chapter 4], and disappearance of some species from EEZs [@Oremus:2020id]. Exploring the detailed effects of climate change on the distribution of shared species is key to the development of local adaptation methods that can anticipate negative impacts to sustainability. For example, understanding how climate change will modify the proportion of transboundary species shared by neighboring EEZs, the time-frame over which such changes will happen (Chapter 4), and the economic consequence of such effects can inform the development of more anticipatory and climate-resilient international treaties improving fisheries management [@Sumaila:2020tp].

>>>>

I overlaid the known distribution of 938 commercially valuable marine species responsible for an average of 96.5% of global EEZ catches between 2005 and 2014, and 280 EEZs of 198 coastal countries (See Methods). I define a ‘stock’ unit as a species in an EEZ, instead of a genetically or morphological distinct unit [@Nelson:2016kk], due to the lack of such biological information being available for almost all the species included in this study [@Teh:2015gd]. While I acknowledge that species could have multiple stocks within an EEZ, many fisheries within a country or EEZ are managed at the species instead of stock level (or even as groups containing multiple species). For example, shrimp (e.g., *Litopenaeus stylirostris* or *Farfantepenaeus californiensis*) along the Pacific coast of Mexico [@DiarioOficialdelaFederacionDOF:2018wt] and hammerhead shark (*Sphyrna zygaena*) in Peru [@MAP:2017uf] are managed at the species level, yet include multiple populations exploited by each country’s fisheries. Moreover, recent research shows connectivity across fish stocks through larval dispersal [@Ramesh:2019va] and adult migration [@Levin:2018in; @Popova:2019fs; @Dunn:2019ca], although considerable level of uncertainty exists at different life stages [@Kaplan:2016hs; @Archambault:2016ie]. For this analysis, I only considered shared species between neighboring EEZs, rather than the species’ extended distribution (e.g., I did not include the high seas). I rely on multiple data sources including occurrence, distribution models and catch data, and only consider a species to be present in a grid cell if all data sources showed positive occurrence (see Methods).

@@

The previous models combine observational data with a series of biotic and abiotic data to determine the probability that a species will be found in a given space at a given time. However, this does not mean that the species in question will actually be there. While the models do use approaches to double-check species occurrences (e.g., ENM-Nereus uses four different species distribution algorithms and SDM-SAU undertakes validation by means of other models), I used a fourth data set to corroborate the models’ outputs. The Sea Around Us estimates total reconstructed catches - catches based on all publicly available information sources and including discards, as well as unreported and illegal catches that are not included in available FAO data - for each country. Catches are also spatially allocated on a $0.5^{\circ}$ x $0.5^{\circ}$ latitude longitude grid [@Zeller2016]. Roughly, the Sea Around Us method consist of the following steps. First, it takes each country's officially reported catch data (e.g., National, FAO or RFMO). Secondly, it uses literature (e.g., peer review, grey literature) to identify missing components (e.g., species, gears) and sources of alternative information for missing components. It then derives country estimates for missing data and creates time series interpolation. Finally, the estimated and official data are aggregated, making up the total reconstructed catch data (see Pauly *et al* [-@Pauly:2019bf] and Zeller *et al.* [-@Zeller2016] for catch reconstruction and spatial allocation details). I used the Sea Around Us catch reconstruction database from 2005 to 2014 as the fourth dataset to estimate transboundary species and to estimate their catch contribution within EEZs. I selected this time frame to investigate the recent (our time frame includes the last decade of available data) contribution of transboundary fisheries to catches and revenue from fisheries and to reduce the uncertainty embedded in the reconstruction process (see Key Uncertainties). Note that in call cases, I report the average catch from 2005 to 2014.

@@

Rules <- c(
  "(Year of Catch > Year Post Max. Min. & Post Max Min Catch < (Max Catch*0.10)) & (Catch > (Max Catch*0.10) & Catch < (Max Catch*0.50)) or Year of Catch < Year of Max. Catch & Catch <= (Max Catch*0.50) or Year of Max Catch = Last Year of data)", #Rebuilding or #Developing
  "Catch > (Max Catch*0.50)", #Max Expploitation
  "Year of Catch > Year of Max. Catch & (Catch > (Max. Catch*0.10) & Catch < (Max. Catch*0.50) or Catch < Max. Catch*0.10", #Over explpoited or collapsed
  "None of the above rules applied"
)

@@

I have identified four key uncertainties in the method utilized that may affect the estimation of transboundary species. Firstly, as I ran the analysis at the species level due to limited spatial-specific data on species sub populations (stocks), I am not able to identify transboundary stocks within EEZs. While I acknowledge that a species could have multiple stocks within an EEZ, many fisheries within a country or EEZ are managed at the species instead of stock level. Also, recent research suggests strong connectivity between some stocks, even when separated by thousands of kilometers [@Ramesh:2019va] providing additional ecological ground for our analysis [@Popova:2019fs; @Dunn:2019ca]. However, it is important to acknowledge that there is considerable uncertainty associated in determining levels of connectivity across time and space for marine populations at different life stages, from larvae [@Kaplan:2016hs] to adults [@Archambault:2016ie]. This is of special concern, but-not-limited-to [@Archambault:2016ie], highly migratory species whose ranges span multiple jurisdictions and the high seas, such as tunas [@Moore:2020cd], billfishes [@Sepulveda:2019hq] and sharks [@Vandeperre:2014jk] challenging management decisions based on meta-populations [@Moore:2020cd]. Here, I limited the definition of “transboundary” to include species spanning only adjacent countries (e.g., the US and Canada), excluding countries that were separated by another nation (e.g., Canada and Mexico) and/or the high seas (e.g., Canada and Spain). Consequently, our results likely provide a conservative estimate of transboundary species, as I did not cover all marine taxa in the world [@Reygondeau:2019uh] and only analyzed species present in all four data sources. Nevertheless, our results are representative of a substantial proportion of the world's marine catches and revenue from economically important marine fisheries. Secondly, the predicted species' distribution is affected by the uncertainties of the environmental data and models used for such predictions. Structural differences within Earth System Models result in variations of oceanic conditions, which undoubtedly affect the ENM-Nereus. I averaged results from the three models in an effort to capture the structural variation across models. Natural climate variability is a major driver of marine species' distributions, potentially removing a species from an EEZ for long periods (e.g., anchovies and sardines "regimes" in the Eastern Pacific are strongly influenced by water temperature decal oscillations [@Chavez:2003jr]. Thus, a species’ distribution can potentially be reduced or shifted in such a way that it only covers one EEZ until oceanic conditions change again and the associated species' distribution expands. To account for such climate variability, I derived the ENM-Nereus results as an average of oceanic conditions between 1970 and 2000. This is not an issue for the SDM-SAU as it does not directly require environmental variables [@Pauly:2016uz]. Thirdly, I assumed that if the Sea Around Us reconstructed data recorded a species as caught in any given grid cell, then the species was present within that grid cell. While catch data are not exempt of uncertainty, in most cases, differences between the Sea Around Us and the FAO self-reported data are smaller towards the end of the time series [@Pauly:2019bf]. Thus, I limited the catch data reference period in our analyses to between 2005 and 2014, the last ten years of data available. Likewise, the spatial allocation of the catch is subjected to imprecision, mainly produced by differences in the spatial scale of the original data and the method employed by the Sea Around Us. Finally, this study’s results only considered species for which all datasets agreed on presence and had an Area Index of at least 25% (e.g., the species shared distribution was at least 25% in each EEZ). Therefore, again, our approach presents a relatively conservative estimate of the number of transboundary species in the world. Using a smaller Area Index will result in more transboundary species (Figure \@ref(fig:ffvFigureS2)).

@@

```{r ffvFigure2, eval = T, fig.width=6, fig.height=4, echo=FALSE, fig.pos= "H", fig.cap="Weighted benefits of transboundary species by km U+00Bx2 and UN sub-region. a) Revenue in thousand 2019 USD. b) Catch in tonnes. Points = sub-region mean \u00B1 s.d. by country."}

@@

I determined the catch trend of each species within each EEZ, classifying them as increasing (Category A), constant (Category B) or decreasing (Category C) (Figures \@ref(fig:ffvFigure3); \@ref(fig:ffvFigureS2)). While previous studies have demonstrated that catch trends may be used to infer whether a stock is healthy, re-building, over-exploited or collapsed [@Kleisner:2011wn], several factors can influence stock status. Our intention here is to examine where the catch trends of transboundary species differ from non-transboundary species between 2005 and 2014 relative to historic catch since 1951 (see Methods). I found significant differences for all catch trend categories for transboundary species (one-way ANOVA, DF(2,459) = 47.94, p < 0.001, $\alpha$ = 0.05; Tukey's post hoc test p $\leq$ 0.001), and no significant differences in catch trend categories for discrete species (one-way ANOVA, DF(2,106) = 1.885, p = 0.157; $\alpha$ = 0.05). I also found significant differences in catch trends when directly comparing transboundary to discrete species categories (MANOVA, DF(2,459) = 19.001, p < 0.001). Overall, transboundary species only targeted by one country are generally less likely to have a decreasing catch trend compared to instances where the shared species is fished by neighboring countries (Table \@ref(tab:ffvTableS2)).

@@

Transboundary fisheries are important to a number of countries with notorious fisheries-related conflicts, including Canada, the USA, the European Union (EU) and Russia [@Spijkers:2019cz]. For example, since 2007, the EU, Norway, Iceland, and the Faroe Islands (Denmark) have been at odds over the size and relative allocation of the total allowable catch for Atlantic mackerel (*Scomber scombrus*) due to the species' climate-driven shift in distribution [@Spijkers:2017ij]. Atlantic Mackerel is a transboundary species that straddles into the high seas. Among the countries involved in the 2007 fisheries dispute, Atlantic mackerel contributed an annual average (between 2005 and 2014) of 598 thousand tonnes (8%) in total catch and USD 850 million (7%) in total fishing revenue.

Climate change is expected to continue changing the distribution and shared proportion of fish stocks among countries, resulting in the emergence of new transboundary species [@Pinsky:2018cb, Chapter 4], and disappearance of some species from EEZs [@Oremus:2020id]. Exploring the detailed effects of climate change on the distribution of shared species is key to the development of local adaptation methods that can anticipate negative impacts to sustainability. For example, understanding how climate change will modify the proportion of transboundary species shared by neighboring EEZs, the time-frame over which such changes will happen (Chapter 4), and the economic consequence of such effects can inform the development of more anticipatory and climate-resilient international treaties improving fisheries management [@Sumaila:2020tp].
